Visualizing Independence Using Extended Association Plots

### 2003-05-28 trunk/vcd/paper/dsc2003.Rnw
small fixes.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@263 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables. 
Now, when the $chi^2$ test statistic turns out to be significant for some data, 
it seems natural to go back to its components, i.e. the  residuals, for a more detailed analysis.

@@

Another example is given in Figure \ref{fig:assoc-classic2}:
this time, we cross-tabulate the variables `Eye' and `Sex'.
The big tiles for brown Eyes suggest convey the impression of an asymmetry between
male and female students, and hence that hypothesis of independence is
to be rejected. However, the $\chi^2$ test statistic is not significant at
a 5\% level: in this example, its value is 7.28, and the corresponding
$p$ value approximately $0.063$. Apparently, the relative size of the
tiles is insufficient to assess the ``strength'' of association
between two variables; again, information on the absolute size of the
residuals would be useful.
Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
that \emph{no} residual is bigger than 2 (or smaller than -2 for
negative ones, respectively) 
because simply no tile is shaded. In fact, as we can depict from
the legend, the overall range of the residuals is approximately $[-1.5,1.5]$.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-classic2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
base::assocplot(margin.table(HairEyeColor, 2:3))
@
\caption{Association plot for `Eye' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-classic2}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(t(margin.table(HairEyeColor, 2:3)), gp = gp.shading)
@
\caption{Extended association plot for `Eye' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-improved2}
\end{center}
\end{figure}

@@

A quick overview of the mutual independence structure of a multi-way
table (i.e., all pairwise \textit{marginal} associations)
can be given by a matrix of plots where each cell contains an
association plot of the corresponding row and column variables. Such an
\emph{association pairsplot} allows us to
quickly scan the association plots for all variable combinations. As
an example, we again use the HairEyeColor data (see Figure
\ref{fig:HEC-pairs}): only the cells corresponding to the variables
`Hair' and `Eye' contain an ``interesting'' association plot with
colored residuals.

@@

The current implementation of association plots in \textsf{R} suffers
from two main disadvantages: First, it is not easy to recycle the
plots in conditioning plots or pairs plots as
they have been implemented using \textsf{R}'s base graphics engine where in
general plotting to relative coordinates is not supported.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some minor advantages and convenient
improvements. Second, the graphics parameters of the rectangles in the
association plots (like in almost all standard \textsf{R}
plots), such as color and line type, cannot be specified for each cell
by the user. To overcome this, the current implementation in \texttt{vcd}
allows the user to specify either arrays of graphics parameters
of the same dimensionality as the object being plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for shading schemes
like those described in the previous sections.

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables using association plots. The
extensions aim at improving the visualization by displaying
the size of the residuals, and by extending the concept of association
plots to multi-way tables. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.

Our work is still not at its end, though: for example, the plots for
multi-way tables currently do not use the same scale for all subplots,
and thus residuals of different subplots are not
comparable. Furthermore, the current color scheme only allows the
detection of ``patterns'' of independence. It would be nice, however, 
that the shading levels were
chosen such that a true visual test of independence can be performed,
that is, such that the presence of ``significant'' residuals lead to a
rejection of the hypothesis of independence. Current work includes the
development of such color schemes, based on an alternative test
statistic---the maximum of the absolute values of the pearson residuals.

>>>>

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables. 
Now, when the $\chi^2$ test statistic turns out to be significant for some data, 
it seems natural to go back to its components, i.e. the  residuals, for a more detailed analysis.

@@

Another example is given in Figure \ref{fig:assoc-classic2}:
this time, we cross-tabulate the variables `Eye' and `Sex'.
For example, the big tiles for blond hair convey the impression of an asymmetry between
male and female students, and hence that hypothesis of independence is
to be rejected. However, the $\chi^2$ test statistic is not significant at
a 5\% level: in this example, its value is 6.221, and the corresponding
$p$ value approximately $0.1013$. Apparently, the relative size of the
tiles is insufficient to assess the ``strength'' of association
between two variables; again, information on the absolute size of the
residuals would be useful.
Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
that \emph{no} residual is bigger than 2 (or smaller than -2 for
negative ones, respectively) 
because simply no tile is shaded. In fact, as we can depict from
the legend, the overall range of the residuals is smaller than $[-1.5,1.5]$.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-classic2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
base::assocplot(margin.table(HairEyeColor, c(1,3)))
@
\caption{Association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-classic2}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(t(margin.table(HairEyeColor, c(1,3))), gp = gp.shading)
@
\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-improved2}
\end{center}
\end{figure}

@@

A general approach for visualizing such kind of hypotheses are
extended mosaic plots. In this section, we give two examples for the
applicability of association plots to multi-way tables.
A quick overview of the mutual independence structure of a multi-way
table (i.e., all pairwise \textit{marginal} associations)
can be given by a matrix of plots where each cell contains an
association plot of the corresponding row and column variables. Such an
\emph{association pairsplot} allows us to
quickly scan the association plots for all variable combinations. As
an example, we again use the HairEyeColor data (see Figure
\ref{fig:HEC-pairs}): only the cells corresponding to the variables
`Hair' and `Eye' contain an ``interesting'' association plot with
colored residuals.

@@

The current implementation of association plots in \textsf{R} suffers
from two main disadvantages: First, it is not easy to recycle the
plots in conditioning plots or pairs plots as
they have been implemented using \textsf{R}'s base graphics engine where in
general plotting to relative coordinates is not supported.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some minor advantages and convenient
improvements. Second, the graphics parameters of the rectangles in the
association plots, such as color and line type, cannot be specified for each cell
by the user. To overcome this, the current implementation in \texttt{vcd}
allows the user to specify either arrays of graphics parameters
of the same dimensionality as the object being plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for shading schemes
like those described in the previous sections.

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables using association plots. The
extensions aim at improving the visualization by displaying
the size of the residuals, and by extending the concept of association
plots to multi-way tables. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.
Our work is still not at its end, though: for example, the plots for
multi-way tables currently do not use the same scale for all subplots,
and thus residuals of different subplots are not
comparable. Furthermore, the current color scheme only allows the
detection of ``patterns'' of independence. It would be nice, however, 
that the shading levels were
chosen such that a true visual test of independence can be performed,
that is, such that the presence of ``significant'' residuals lead to a
rejection of the hypothesis of independence. Current work includes the
development of such color schemes, based on an alternative test
statistic---the maximum of the absolute values of the pearson residuals.
